Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-02-27 NovelLBM.tex
mark todos

\begin{table}[H]
	\centering
		\begin{tabular}{l|p{3cm}|p{1.6cm}|p{1.6cm}|p{1.6cm}|p{2cm}}
			Author & Material & $\eta_0/\unit{\pascal\second}$ & $\eta_\infty/\unit{\pascal\second}$ & $R_\eta$ & extracted from \\
			\hline
			Amorim\cite{amorim2021} & pre-crosslinked alginate &  $>\num{1e4}$ & $<\num{3e-1}$ & $>\num{3e4}$ & Fig. 3b \\
			\rowcolor{secondaryBackground}
			Pössl\cite{possl2021} & blend & \num{2e2} & \num{5e-1} & \num{4e2} & Figure 4 \\
%			Paxton\cite{paxton2017} &Nivea Creme\n Alginate& 1e5\n5e4 & <1\n<5e1 & >1e5\n>1e3 & Fig. 4b \\
			Paxton\cite{paxton2017} & Alginate& \num{5e4} & $<\num{5e1}$ & $>\num{1e3}$ & Fig. 4b \\
			\rowcolor{secondaryBackground}
			O'Connell\cite{oconnell2020} & pluronic F-127 & $>\num{1e6}$ & $<\num{8}$ & $>\num{2e5}$ & Chapter 7 Fig. 13 \\
		\end{tabular}
	\caption{Compilation of literature data on typical bioinks showing that a very viscosity ratio $ R_ \eta $ is a generic feature of these liquids.
	\SG{ cross-check these values }\RK{Done and formatted}
	}
	\label{tab:bioinks}
\end{table}

@@

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
For this, ... \SG{ more details }
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in planar channel flow.
		The time evolution of the center-line velocity is shown in the inset \ref{fig:2DTimeEvolution}\SG{ to do }\RK{Not an inset but plotted. Has to be inset?}.
		(b) Velocity profile for methyl cellulose.
}
		\label{fig:2Dvel}
	\end{figure}

We proceed to evaluate the viscosity shuffling method for a cylindrical Poiseuille flow.
The semi-analytical solution for this situation can be derived similarly as for the planar case and is given in appendix~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical situation is slightly more complex due to discretization errors ("staircase effect") when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameters as above, but double the pressure gradient to \SI{-2e7}{\pascal\per\meter} and approximately \SI{-2.5e7}{\pascal\per\meter}, respectively, thereby mirroring the relation in Newtonian Poiseuille flow where twice the pressure is required to obtain the same flow rate in cylindrical compared to planar gemeotries.
The channel has a diameter of \SI{20}{\micro\meter} and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure~\ref{fig:3Dvel} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014} and \num{0.06}, respectively.
The spatially resolved error in figure~\ref{fig:3DvelErr} of the appendix shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.
The robustness against pressure gradient changes is shown in figure~\ref{fig:pressureErr}~(b) of the appendix.

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in cylindrical channel flow.
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset \ref{fig:3DTimeEvolution}.\RK{currently not an inset}
		(b) Velocity profile for methyl cellulose.
				}
		\label{fig:3Dvel}
	\end{figure}

@@

As can be seen in figure~\ref{fig:malaspinas2010}, the accuracy of our method does not depend on the Weissenberg Number and is generally somewhat better than the method of \cite{malaspinas2010}.
We note a certain dependence of the accuracy on $R_\nu$ which we consider nevertheless acceptable given the overall smallness of the error.
Finally, we find that the error of our method decreases with lattice resolution to the power of \num{-2}. \RK{They see diminishing returns we don't} \SG{ explain }

@@

The first application stems from bioprinting where the bioink is pushed through a conical nozzle \cite{possl2021, Poologasundarampillai.2021, oconnell2022}.
The nozzle radius is \SI{1.8}{\milli\meter} at the inlet and \SI{200}{\micro\meter} at the outlet.
The nozzle has a length of \SI{18}{\milli\meter}.
Due to the difference in inlet/outlet radii, the system cannot be modeled with periodic boundary conditions, and we use inflow/outflow boundary conditions as described in section~\ref{sec:Methods} instead.
The required values for velocity and polymer conformation are obtained from simulations of pipes of the respective radii and a fixed volume flow of $\approx\SI{3}{\micro\liter\per\second}$.
The resulting flow field can be seen in figure \ref{fig:Nozzle}~(a).
To quantify the total stress as a single number, we use the von Mises stress which is defined as \cite{Mueller.2023}:
\begin{align}
	\sigma_\text{vM} = \sqrt{\frac{1}{2}\left[\left(\sigma_1 - \sigma_2\right)^2 + \left(\sigma_2 - \sigma_3\right)^2 + \left(\sigma_3 - \sigma_1\right)^2\right]}.
\end{align}
It can be seen that the highest stresses arise at the nozzle exit in direct vicinity to the wall.
This means that even for fast flows, cells flowing close to the center experience less stress than cells flowing near the walls in agreement with earlier results in Newtonian fluids \cite{Mueller.2023}.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_v.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_sigma_vM.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_tau_12.pdf}
		\caption{Velocity profile (a), von Mises stress (b) and shear stress (c) for methyl cellulose in a conical nozzle as used for bioprinting. For the calculated viscosity see figure~\ref{fig:Nozzle/RT-DC_eta}.}
		\label{fig:Nozzle}
	\end{figure}

@@

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
Our derivation is to the one in Oliveira \cite{oliveira1999}. For both Poiseuille flow and shear flow, we assume that the velocity is unidirectional: 
\begin{equation}
	\vec{u} = \dot{\gamma}\left(y\right)\hat{e}_x.
\end{equation}
\SG{ This equation does not work, $ \dot{ \gamma  }$ is a shear rate, it should read $ \partial u_x / \partial y = \dot{ \gamma  }$. }

@@

In axisymmetric flows $y$ in eqs.~(\ref{eq:PTTDerivation1}, \ref{eq:PTTDerivation2}) needs to be replaced by the radial coordinate $r$. \SG{ no factor 2? }
With $\Tr\ma{\tau} = \tau_{xy}$. \SG{ should be $ \tau _{ xx }$? }

@@

>>>>

\begin{table}[H]
	\centering
		\begin{tabular}{l|p{3cm}|p{1.8cm}|p{1.8cm}|p{1.8cm}|p{2cm}}
			Author & Material & $\eta_0/\unit{\pascal\second}$ & $\eta_\infty/\unit{\pascal\second}$ & $R_\eta$ & extracted from \\
			\hline
			Amorim\cite{amorim2021} & pre-crosslinked alginate &  $>\num{1e4}$ & $<\num{3e-1}$ & $>\num{3e4}$ & Fig. 3b \\
			\rowcolor{secondaryBackground}
			Pössl\cite{possl2021} & blend & \num{2e2} & \num{5e-1} & \num{4e2} & Figure 4 \\
%			Paxton\cite{paxton2017} &Nivea Creme\n Alginate& 1e5\n5e4 & <1\n<5e1 & >1e5\n>1e3 & Fig. 4b \\
			Paxton\cite{paxton2017} & Alginate& \num{5e4} & $<\num{5e1}$ & $>\num{1e3}$ & Fig. 4b \\
			\rowcolor{secondaryBackground}
			Jalaal\cite{jalaal2017} & pluronic F-127 & $>\num{1e3}$ & $<\num{1e-1}$ & $>\num{1e4}$ & FIG. 6. (e) \\
		\end{tabular}
	\caption{Compilation of literature data on typical bioinks showing that a very viscosity ratio $ R_ \eta $ is a generic feature of these liquids.
	\SG{ cross-check these values }\RK{Done and formatted}
	}
	\label{tab:bioinks}
\end{table}

@@

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
For this, ... \SG{ more details }\todo{SRT collision operator as implemented, advection as implemented, the whole as implemented appendix that was deleted}
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in planar channel flow.
		The time evolution of the center-line velocity is shown in the inset \ref{fig:2DTimeEvolution}\SG{ to do }\RK{Not an inset but plotted. Has to be inset?}.
		(b) Velocity profile for methyl cellulose.}
		\label{fig:2Dvel}
	\end{figure}

We proceed to evaluate the viscosity shuffling method for a cylindrical Poiseuille flow.
The semi-analytical solution for this situation can be derived similarly as for the planar case and is given in appendix~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical situation is slightly more complex due to discretization errors ("staircase effect") when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameters as above, but double the pressure gradient to \SI{-2e7}{\pascal\per\meter} and approximately \SI{-2.5e7}{\pascal\per\meter}, respectively, thereby mirroring the relation in Newtonian Poiseuille flow where twice the pressure is required to obtain the same flow rate in cylindrical compared to planar geometries.
The channel has a diameter of \SI{20}{\micro\meter} and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure~\ref{fig:3Dvel} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014} and \num{0.06}, respectively.
The spatially resolved error in figure~\ref{fig:3DvelErr} of the appendix shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.
The robustness against pressure gradient changes is shown in figure~\ref{fig:pressureErr}~(b) of the appendix.

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in cylindrical channel flow.
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset \ref{fig:3DTimeEvolution}.\RK{currently not an inset}
		(b) Velocity profile for methyl cellulose.}
		\label{fig:3Dvel}
	\end{figure}

@@

As can be seen in figure~\ref{fig:malaspinas2010}, the accuracy of our method does not depend on the Weissenberg Number and is generally somewhat better than the method of \cite{malaspinas2010}.
We note a certain dependence of the accuracy on $R_\nu$ which we consider nevertheless acceptable given the overall smallness of the error.
Finally, we find that the error of our method decreases with lattice resolution to the power of \num{-2}. \RK{They see diminishing returns we don't} \SG{ explain }\RK{The topmost line starts from the first green dot and also is a power of -2 line, the second point is already a little of and the third and fourth deviate more and more. One could try to do a line with a smaller power, however that does not quite fit the data either. This is particularly evident for the red points. It looks like Malaspinas has about power of -2 for small N but the slope decreases for higher N. This can be explained through other error terms (for example their artificial viscosity) becoming more relevant as the lattice resolution caused error decreases. As we do not have those our error continues to decrease as expected}

@@

The first application stems from bioprinting where the bioink is pushed through a conical nozzle \cite{possl2021, Poologasundarampillai.2021, oconnell2022}.
The nozzle radius is \SI{1.8}{\milli\meter} at the inlet and \SI{200}{\micro\meter} at the outlet.
The nozzle has a length of \SI{18}{\milli\meter}.
Due to the difference in inlet/outlet radii, the system cannot be modeled with periodic boundary conditions, and we use inflow/outflow boundary conditions as described in section~\ref{sec:Methods} instead.
The required values for velocity and polymer conformation are obtained from simulations of pipes of the respective radii and a fixed volume flow of $\approx\SI{3}{\micro\liter\per\second}$.
The resulting flow field can be seen in figure \ref{fig:Nozzle}~(a).
To quantify the total stress as a single number, we use the von Mises stress which is defined as \cite{Mueller.2023}:
\begin{align}
	\sigma_\text{vM} = \sqrt{\frac{1}{2}\left[\left(\sigma_1 - \sigma_2\right)^2 + \left(\sigma_2 - \sigma_3\right)^2 + \left(\sigma_3 - \sigma_1\right)^2\right]}.
\end{align}
It can be seen that the highest stresses arise at the nozzle exit in direct vicinity to the wall.
This means that even for fast flows, cells flowing close to the center experience less stress than cells flowing near the walls in agreement with earlier results in Newtonian fluids \cite{Mueller.2023}.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_v.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_sigma_vM.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_tau_12.pdf}
		\caption{Velocity profile (a), von Mises stress (b) and shear stress (c) for methyl cellulose in a conical nozzle as used for bioprinting. For the calculated viscosity see figure~\ref{fig:Nozzle/RT-DC_eta}.\RK{The black lines currently hide our resolution shall I remove them to be consistent with the RT-DC plots or do we prefer this?}}
		\label{fig:Nozzle}
	\end{figure}

@@

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
Our derivation is \RK{hier könnte ein Adjektiv stehen} to the one in Oliveira \cite{oliveira1999}. For both Poiseuille flow and shear flow, we assume that the velocity is unidirectional:
\begin{equation}
	\vec{u} = u_x\left(y\right)\hat{e}_x
\end{equation}
\begin{equation}
	\dot{\gamma}\left(y\right) = \frac{\partial u_x}{\partial y}
\end{equation}
\SG{ This equation does not work, $ \dot{ \gamma  }$ is a shear rate, it should read $ \partial u_x / \partial y = \dot{ \gamma  }$. }\RK{done, was supposed to read g(y) but could not read my own handwriting when copying... Made it a little more clear}

@@

In axisymmetric flows $y$ in eqs.~(\ref{eq:PTTDerivation1}, \ref{eq:PTTDerivation2}) needs to be replaced by the radial coordinate $r$. \RK{2D case is also axisymmetric?}\SG{ no factor 2? }\RK{No, factor 2 comes from the NS eq. and thus is only relevant for Poiseuille solution. This replacement is also much more relevant there as here all ys disappear}
With $\Tr\ma{\tau} = \tau_{xx}$. \SG{ should be $ \tau _{ xx }$? }\RK{yes}

@@

	\begin{figure}[H]
		\centering
		%\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/ShuffleInPoiseuille.pdf}
		\caption{In Poiseuille flow, the result is near identical for different $\alpha_\text{s}$. The error reveals minor differences. Aside from the wall nodes, the largest error appears for small shuffle fractions. For large ones, the error appears to become independent of $\alpha_\text{s}$. This indicates, that the observed difference might be due to the Mach number changing for different $\alpha_\text{s}$ and not due to the shuffling itself.\RK{I could probably prove this by running sims at different $\alpha_s$ that consistently are at Ma << 1}}
		\label{fig:ShuffleInPoiseuille}
	\end{figure}\todo{include plot, reference in main text}
